Software Citation Principles

### 2016-03-03 sc-principles/software-citation-principles.tex
Add single space before \cite command and minor edits

\katznote{to add cite of Matthew Mayernik's similar work} 
Software and other digital resources currently appear in publications in very inconsistent ways.
For example, a random sample of 90 articles in the biology literature found seven different ways that software was mentioned, including simple names in the full-text, URLs in footnote, and different kinds of mentions in references lists: project names or websites, user manuals, publications that describe or introduce the software~\cite{howison2015jasist}. Table \ref{tab:mentions} shows examples of these varied forms of software mentions and the frequency with which they were encountered. 
%Even in the top journals (top 10 by impact factor), only 38\% of mentions included an entry of some sort in the reference list and fully 36\% were informal, following no recognizable style or format. 
Many of these kinds of mentions fail to perform the functions needed of citations, and their very diversity and frequent informality undermines the integration of software work into bibliometrics and other analyses. 
In some ways the situation is not too surprising: practices are diverse and informal because authors receive little guidance from editors, style guides, and their scientific communities, as well as weak support from citation software (such as Endnote and Zotero).
% In Howison and Bullard's study only 24\% of journals provided any guidance and software is treated inconsistently by style guides, if it appears at all.

\rowcolors{2}{white}{gray!25}
\begin{table}[htbp]
\caption{Varieties of software mentions in publications, from  \cite{howison2015jasist}} 
\centering
\scriptsize\setlength{\tabcolsep}{2.5pt}
\begin{tabular}{@{}l l l } % p{0.6\textwidth}@{}}
\toprule
Mention Type & \multicolumn{2}{c}{Count (n=286)} \\ % & Example \\
\midrule
Cite to publication & 105 & 37\% \\ %& \ldots was calculated using biosys (Swofford \& Selander 1981). \newline Ref: Swofford DL, Selander RB (1981) biosys-1: a Fortran program for the comprehensive analyses of electrophoretic data in population genetics and systematics. \emph{Journal of Heredity}, 72, 281–283.\\
Cite to users manual & 6 & 2\% \\ %& \ldots as analyzed by the BIAevaluation software (Biacore, 1997). \newline Ref: Biacore, I. (1997). BIAevaluation Software Handbook, version 3.0 (Uppsala, Sweden: Biacore, Inc) \\
Cite to name or website & 15 & 5\% \\ %& \ldots using the program Autodecay version 4.0.29 PPC (Eriksson 1998). \newline Ref: ERIKSSON, T. 1998. Autodecay, vers. 4.0.29 Stockholm:Department of Botany. \\
Instrument-like & 53 & 19\% \\ %& \ldots calculated by t-test using the Prism 3.0 software (GraphPad Software, San Diego, CA, USA). \\
URL in text & 13 & 5\% \\ %& \ldots freely available from http://www.cibiv.at/software/pda/. \\
In-text name only & 90 & 31\% \\ %& \ldots were analyzed using MapQTL (4.0) software. \\
Not even name & 4 & 1\% \\ %& \ldots was carried out using software implemented in the Java programming language. \\
\bottomrule
\end{tabular}
\label{tab:mentions}
\end{table}%

@@

We have documented and analyzed a set of use cases related to software citation in \cite{SC-Use-Cases}.
Table~\ref{tab:use_cases} summarizes these use cases and makes clear what the requirements are for software citation in each case.
\katznote{not sure if this next long sentence should be here, should be in the table caption, or should be footnotes to the table, or even more than one of these}
\niemnote{Not sure either; definitely should include in the text due to length. Maybe an abbreviated version in the caption?}\flnote{If these ``definitions'' are not used anywhere else than in the table (and they don't really seem to be so far), then I tend towards putting this in the caption. That's where they are needed (in the table). I don't like footnotes for something like this.}\dmnote{Agree it would be better in the caption.}
In the table,
``researcher'' includes both academic researchers (e.g., postdoc, tenure-track faculty member) and research software engineers;
``publisher'' includes both traditional publishers that publish text and\slash or software papers as well as archives such as Zenodo that directly publish software;
``reproduce'' can mean reproduction, replication, verification, validation, repeatability, and\slash or utility;
examples of indexers include Scopus, Web of Science, Google Scholar, and Microsoft Academic Search;
domain groups include bioCADDIE, CIG, etc.;\dmnote{Not sure what CIG stands for, and acronyms should probably be explained or linked}
``repository'' refers to open-source software repositories such as ASCL, GitHub, Netlib, CRAN, institutional repositories;
``funder'' is a group that funds software or work using software; and
``UID'' means unique ID, such as a DOI or PURL.

@@

AAS workshop (\cite{aas-software-index}), CRediT project \cite{casrai-credit}, Ontosoft \cite{ontosoft}

@@

The following set of software citation principles were originally based on an adaptation of the parallel set of data citation principles~\cite{data-citation-principles}, and then were modified based on working group discussions and the information from the use cases and the related work.

\begin{enumerate}
\item \textbf{Importance}:
Software should be considered a legitimate, citable product of research. Software citations should be accorded the same importance in the scholarly record as citations of other research objects, such as publications and data. Software should be cited whenever and wherever a research product (such as a paper) relies upon it, specifically, as part of the standard reference list for that research product.
\item \textbf{Credit and Attribution}:
Software citations should facilitate giving scholarly credit and normative and legal attribution to all contributors to the software, recognizing that a single style or mechanism of attribution may not be applicable to all software.\scnote{At least four styles (AMS, APA, Chicago, MLA) for software citation are illustrated with examples in ~\cite{lipson2011cite}.}
\item \textbf{Unique Identification}:
A software citation should include a method for identification that is machine actionable, globally unique, interoperable, and widely used by a community.
\item \textbf{Persistence}:
Unique identifiers, and metadata describing the software, and its disposition, should persist -- even beyond the lifespan of the software they describe.
\item \textbf{Accessibility}:
Software citations should facilitate access to the software themselves and to such associated metadata, documentation, data, and other materials, as are necessary for both humans and machines to make informed use of the referenced software.\dmnote{Should provide a location where the software is available. Should we say anything about environments?}
\item \textbf{Specificity}:
Software citations should facilitate identification of and access to the specific version of software that was used.  Software identification should be as specific as necessary, such as using version numbers, revision numbers, or variants such as platforms.
%\item \textbf{Interoperability and Flexibility}:
%Software citation methods should be sufficiently flexible to accommodate the variant practices among communities, but should not differ so much that they compromise interoperability of software citation practices across communities.
\end{enumerate}
\dmnote{ Anything missing? Configurations? Interactive software? Screenshots? \ldots ?}

@@

Sou-Cheng Choi, NORC at the University of Chicago \& Illinois Institute of Technology

@@

James Howison, University of Texas

Lorraine Hwang, University of California,  Davis

@@

Daniel S. Katz, University of Chicago \& Argonne National Laboratory (co-chair)

@@

Brian Matthews, Science and Technology Facilities Council

@@

Soren Scott, Independent Researcher

@@

Ilian Todorov, Science and Technology Facilities Council

>>>>

\katznote{add cites to James's study how what people cite software and Matthew Mayernik's similar one as part of motivation}

@@

We have documented and analyzed a set of use cases related to software citation in \cite{SC-Use-Cases}.
Table~\ref{tab:use_cases} summarizes these use cases and makes clear what the requirements are for software citation in each case.
\katznote{not sure if this next long sentence should be here, should be in the table caption, or should be footnotes to the table, or even more than one of these}
\niemnote{Not sure either; definitely should include in the text due to length. Maybe an abbreviated version in the caption?}\flnote{If these ``definitions'' are not used anywhere else than in the table (and they don't really seem to be so far), then I tend towards putting this in the caption. That's where they are needed (in the table). I don't like footnotes for something like this.}\dmnote{Agree it would be better in the caption.}
In the table,
``researcher'' includes both academic researchers (e.g., postdoc, tenure-track faculty member) and research software engineers;
``publisher'' includes both traditional publishers that publish text and\slash or software papers as well as archives such as Zenodo that directly publish software;
``reproduce'' can mean reproduction, replication, verification, validation, repeatability, and\slash or utility;
examples of indexers include Scopus, Web of Science, Google Scholar, and Microsoft Academic Search;
domain groups include bioCADDIE, CIG, etc.;\dmnote{Not sure what CIG stands for, and acronyms should probably be explained or linked}
``repository'' refers to public software repositories such as ASCL,
GitHub, Netlib, CRAN, institutional repositories;
``funder'' is a group that funds software or work using software; and
``UID'' means unique ID, such as a DOI or PURL.

@@

AAS workshop \cite{aas-software-index}, CRediT project \cite{casrai-credit}, Ontosoft \cite{ontosoft}

@@

\katznote{The current content of this section is liberally borrowed and loosely adapted from \cite{data-citation-principles}.  We we will want to add some text here explaining that we started with that document, and then modified it to take into account the information from the use cases and the related work, once we do so.}

\begin{enumerate}
\item \textbf{Importance}:
Software should be considered a legitimate, citable product of research. Software citations should be accorded the same importance in the scholarly record as citations of other research objects, such as publications and data. Software should be cited whenever and wherever a research product (such as a paper) relies upon it, specifically, as part of the standard reference list for that research product.
\item \textbf{Credit and Attribution}
Software citations should facilitate giving scholarly credit and normative
and legal attribution to all contributors to the software, recognizing
that a single style or mechanism of attribution may not be applicable to
all software.\scnote{At least four styles (AMS, APA, Chicago, MLA) for
software citation are illustrated with examples in~\cite{lipson2011cite}.}
\item \textbf{Unique Identification}
A software citation should include a persistent method for identification that is machine actionable, globally unique, and widely used by a community.
\item \textbf{Accessibility}
Software citations should facilitate access to the software themselves and
to such associated metadata, documentation, data, and other materials, as
are necessary for both humans and machines to make informed use of the
referenced software.
\item \textbf{Persistence}
Unique identifiers, and metadata describing the software, and its disposition, should persist -- even beyond the lifespan of the software they describe.
\item \textbf{Specificity and Verifiability}
Software citations should facilitate identification of, access to, and verification of the specific software that support a claim. Citations or citation metadata should include information about provenance and fixity sufficient to facilitate verifying that the specific version and/or portion of software retrieved subsequently is the same as was originally cited.
\item \textbf{Interoperability and Flexibility}
Software citation methods should be sufficiently flexible to accommodate the variant practices among communities, but should not differ so much that they compromise interoperability of software citation practices across communities.
\end{enumerate}

@@

Sou-Cheng Choi , NORC at the University of Chicago \& Illinois Institute of Technology

@@

James Howison, UTexas

Lorraine Hwang, UC Davis

@@

Daniel S. Katz, U Chicago \& Argonne Natl Lab (co-chair)

@@

Brian Matthews, Science and Technology Facilites Council

@@

Soren Scott , Independent Researcher

@@

Ilian Todorov, STFC
